Palomar 5 Gaps

### 2024-09-20 paper/main.tex
tightened up the methods

    Briefly, the x'coordinate represents the position of a particle along the orbit relative to the globular cluster—whether it is ahead or behind the cluster. The y' coordinate measures the particle's distance within the orbital plane, where positive values indicate that the particle is farther from the Galactic center and negative values indicate that it is closer.

@@

      With the smoothed 1D density profiles in hand, we treat the \textit{vanilla} simulations as the reference model and search for regions in the \textit{full} simulations that are significantly underdense compared to \textit{vanilla}, surpassing stochastic fluctuations. To accomplish this, we impose a signal-to-noise ratio threshold, $\mathcal{SNR}$. The signal is defined as the log of the counts per bin from the \textit{vanilla} 1D density profile, and errors are propagated assuming a Poisson distribution. We then compute a threshold for the number of counts in the \textit{vanilla} simulation, $N_v$, using the transcendental equation:
      \begin{equation}
          \mathcal{SNR} = \ln(10) \log_{10}\left(N_v\right) \sqrt{N_v}.
        \end{equation}
      By setting $\mathcal{SNR} = 5$, we solve for $N$ using \texttt{scipy.optimize.fsolve}, finding that $N$ must be greater than 7. After discarding insignificant bins (i.e., those with counts below the threshold), we compute the log ratio of the counts between the \textit{vanilla} and full simulations:
      \begin{equation}
          \mathcal{R}_i = \log_{10}\left(\frac{N_{f,i}}{N_{v,i}}\right),
        \end{equation}
      where $\mathcal{R}_i$ is the log ratio, $N_{f,i}$ are the counts from the full simulation, and $N_{v,i}$ are the counts from the \textit{vanilla} simulation for each bin $i$. We then analyze the distribution of $\mathcal{R}_i$. If the differences between the density profiles are primarily due to stochastic processes of similar magnitude, this distribution should resemble a Gaussian as expected from the central limit theorem. Thus, we flag all regions where the density is underdense by more than two standard deviations, which should highlight regions whose underdensity is unlikely to be the result of the sum of stochastic processess but rather the passage of another globular cluster.

@@

      \citet{sanders2016dynamics} extended the analysis of Erkal et al. (2015), demonstrating that action-angle variables provide a useful coordinate system for analyzing stream evolution, as actions are conserved and the angles associated with the stream's growth evolve linearly over time. Although we became aware of this work only after completing our analysis, we note that $\tau$ is a suitable approximation and behaves similarly to the angle variable corresponding to the azimuthal action: $\tau \approx \theta_{\phi,i} - \theta_{\phi,\text{GC}}$

      The core of our analysis is presented in Fig.~\ref{fig:force-on-orbit}. The bottom panel shows the evolution of the stream density over time. To avoid extreme low-density regions at the stream's edges, we applied a density threshold to focus on the more significant areas of the stream. Next, we modeled Palomar 5's orbit as a proxy for its stream and sampled points along the orbit to measure the gravitational force exerted by other globular clusters. This force data is displayed in the top panel of Fig.~\ref{fig:force-on-orbit}, showing how the total gravitational acceleration on Palomar 5's stream evolves over its length throughout the simulation.

@@

    From this analysis, we can conclude that a more massive perturber, passing closer to the particle, and moving more slowly, will have a greater impact. It's important to note that the momentum change is inversely proportional to the velocity of the perturber, which contrasts with the intuition from elastic collisions, such as those between billiard balls, where higher velocities generally result in greater impact.

    Erkal et al. (2015) extended this impact theory to a one-dimensional stream, studying how the momentum change of a particle varies as a function of its distance from the point of impact. In their analysis, the perturber was modeled not as a point mass, but as a Plummer sphere with a characteristic length scale $r_p$. Since the stream is not zero-dimensional and has length, both the parallel and perpendicular components of the velocity must be considered. Consequently, five parameters determine the change in velocity of a given particle: $M$, $r_p$, $b$, $W_\parallel$, and $W_\perp$. Our goal is to constrain these parameters during the impact in our simulations and compare them to close approaches that do not generate gaps, in order to establish the criteria necessary for gap formation.

@@

    In this refinement process, we aim to pinpoint both the exact location of the impact along the stream and the precise moment it occurred. To do so, we fit a third-order parametric polynomial to the stream, using the saved snapshots from our simulations.

    \begin{equation}
      \vec{s}(\tau) = 
      \left\{
        \begin{aligned}
          x(\tau) &= a_0 + a_1 \tau + a_2 \tau^2 + a_3 \tau^3 \\ 
          y(\tau) &= b_0 + b_1 \tau + b_2 \tau^2 + b_3 \tau^3 \\
          z(\tau) &= c_0 + c_1 \tau + c_2 \tau^2 + c_3 \tau^3
        \end{aligned}
      \right.
      \end{equation}  
    where $x$, $y$, and $z$ represent the parametric line describing the stream in galactocentric coordinates, $\tau$ is the stream coordinate in time as described in the previous section, and is used as the independent variable in this analysis. The coefficients $a_i$, $b_i$, and $c_i$ are the polynomial coefficients. We found that a second-order polynomial was insufficient to capture the curvature along the full length of the stream, with divergence from the star particle positions occurring at the extremes. A third-order polynomial was sufficient and desirable, as it is the lowest order that adequately captures the path of the stream over the entire length under consideration.

@@

    As a result, each polynomial coefficient can now be expressed as a function of time. Consequently, we can parameterize the stream as a function of both simulation time and position along the stream:

    \begin{equation}
      \vec{s}(t,\tau) = 
      \left\{
      \begin{aligned}
        x(t,\tau) &= a_0(t) + a_1(t)\tau + a_2(t) \tau^2 + a_3(t)\tau^3 \\ 
        y(t,\tau) &= b_0(t) + b_1(t)\tau + b_2(t) \tau^2 + b_3(t)\tau^3 \\
        z(t,\tau) &= c_0(t) + c_1(t)\tau + c_2(t) \tau^2 + c_3(t)\tau^3.
        \end{aligned}
      \right.
    \end{equation}
    The values of the coefficients as a function of time are obtained through linear interpolation, ensuring that the coefficients at the snapshot times match the values constrained by the simulation data.

    Next, we fit the trajectory of the perturber with a second-order polynomial. With equations for both the stream and the perturber as functions of time, we identify the time and location of impact by minimizing a cost function, defined as the distance between the stream and the perturber:

    \begin{equation} b(t, \tau) = \left\lVert \vec{s}(t, \tau) - \vec{p}(t) \right\rVert, \end{equation}

    where $\vec{s}(t, \tau)$ is the galactocentric position of a point on the stream, $\vec{p}(t)$ is the position of the perturber, and $b$ is the distance between the two. The minimum value of $b$, denoted as $\text{min}(b)$, represents the impact parameter. The minimization is carried out using \texttt{scipy}'s optimization package with the \textit{L-BFGS-B} method, which allows us to constrain $t$ and $\tau$, ensuring no extrapolation occurs.

    Once this minimization is performed, determining the relative velocity becomes straightforward. Since the minimization provides the impact parameter, time of impact, and corresponding value of $\tau$, we can compute the derivatives of the parametric equations at $t_{\text{min}}$ and $\tau_{\text{min}}$. The parallel and perpendicular components of the perturber's velocity relative to the stream are given by:

    \begin{equation}
        \begin{aligned}
          \delta \vec{v} &=\vec{v_p} - \vec{v_s} \\
          w_\parallel &= \left(\delta \vec{v}\right)\cdot \hat{v_s}\\  
          w_\perp &=  \sqrt{\Delta v ^2 - w_\parallel ^ 2}
          \end{aligned}
        \end{equation}

    where $\vec{v}_p$ and $\vec{v}_s$ are the velocities of the perturber and the stream, respectively. For each of the 50 simulations, this information was compute for the strongs 5 flybys of a perturber with the stream. Thus, we compute a sample of 250 impacts, and flag those that give way to gaps.

>>>>

    Briefly, the x' coordinate represents the position of a particle along the orbit relative to the globular cluster—whether it is ahead or behind the cluster. The y' coordinate measures the particle's distance within the orbital plane, where positive values indicate that the particle is farther from the Galactic center and negative values indicate that it is closer.

@@

      With the smoothed 1D density profiles in hand, we treat the \textit{vanilla} simulations as the reference model and search for regions in the \textit{full} simulations that are significantly underdense compared to \textit{vanilla}, surpassing stochastic fluctuations. To accomplish this, we impose a signal-to-noise ratio threshold, $\mathcal{SNR}$. The signal is defined as the log of the counts per bin from the \textit{vanilla} 1D density profile, and errors are propagated assuming a Poisson distribution. We then compute a threshold for the number of counts in the \textit{vanilla} simulation, $N_v$, using the transcendental equation:
      \begin{equation}
          \mathcal{SNR} = \ln(10) \log_{10}\left(N_v\right) \sqrt{N_v}.
        \end{equation} \label{eq:density_threshold}
      By setting $\mathcal{SNR} = 5$, we solve for $N$ using \texttt{scipy.optimize.fsolve}, finding that $N$ must be greater than 7. After discarding insignificant bins (i.e., those with counts below the threshold), we compute the log ratio of the counts between the \textit{vanilla} and full simulations:
      \begin{equation}
          \mathcal{R}_i = \log_{10}\left(\frac{N_{f,i}}{N_{v,i}}\right),
        \end{equation}
      where $\mathcal{R}_i$ is the log ratio, $N_{f,i}$ are the counts from the full simulation, and $N_{v,i}$ are the counts from the \textit{vanilla} simulation for each bin $i$. We then analyze the distribution of $\mathcal{R}_i$. If the differences between the density profiles are primarily due to stochastic processes of similar magnitude, this distribution should resemble a Gaussian as expected from the central limit theorem. Thus, we flag all regions where the density is underdense by more than two standard deviations, which should highlight regions whose underdensity is unlikely to be the result of the sum of stochastic processess but rather the passage of another globular cluster.

@@

      \citet{sanders2016dynamics} extended the analysis of Erkal et al. (2015), demonstrating that action-angle variables provide a useful coordinate system for analyzing stream evolution, as actions are conserved and the angles associated with the stream's growth evolve linearly over time. Although we became aware of this work only after completing our analysis, we note that $\tau$ is a suitable approximation and behaves similarly to the angle variable corresponding to the azimuthal action: $\tau \approx \theta_{\phi,i} - \theta_{\phi,\text{GC}}$.

      The core of our analysis is presented in Fig.~\ref{fig:force-on-orbit}. The bottom panel shows the evolution of the stream density over time. To avoid extreme low-density regions at the stream's edges, we applied the same density threshold as from eq.~\ref{eq:density_threshold} to focus on the more significant areas of the stream. Next, we modeled Palomar 5's orbit as a proxy for its stream and sampled points along the orbit to measure the gravitational force exerted by other globular clusters. This force data is displayed in the top panel of Fig.~\ref{fig:force-on-orbit}, showing how the total gravitational acceleration on Palomar 5's stream evolves over its length throughout the simulation.

@@

    From this analysis, we can conclude that a more massive perturber, passing closer to the particle, and moving more slowly, will have a greater impact. It's important to note that the momentum change is inversely proportional to the velocity of the perturber, which contrasts with the intuition from elastic collisions, such as those between billiard balls, where higher velocities result in greater impacts.

    Erkal et al. (2015) extended this impact theory from one point mass impacting another point mass, to studying how an extended body impacts as stream. This quantifies the change in momentum of a given particle as a function of its distance from the point of greatest impact. In their analysis, the perturber was modeled as a Plummer sphere. Since the stream is not zero-dimensional and has length, both the parallel and perpendicular components of the velocity must be considered. Consequently, five parameters determine the change in velocity of a given particle: $M$, $r_p$, $b$, $W_\parallel$, and $W_\perp$. Our goal is to constrain these parameters during the impact in our simulations and compare them to close approaches that do not generate gaps, in order to establish the criteria necessary for gap formation.

@@

    In this refinement process, we aim to pinpoint both the exact location of the impact along the stream and the precise moment it occurred. To do so, we fit a third-order parametric polynomial to the stream, using the saved snapshots from our simulations:
    \begin{equation}
      \vec{s}(\tau) = 
      \left\{
        \begin{aligned}
          x(\tau) &= a_0 + a_1 \tau + a_2 \tau^2 + a_3 \tau^3 \\ 
          y(\tau) &= b_0 + b_1 \tau + b_2 \tau^2 + b_3 \tau^3 \\
          z(\tau) &= c_0 + c_1 \tau + c_2 \tau^2 + c_3 \tau^3
        \end{aligned}
      \right.
      \end{equation}  
    where $x$, $y$, and $z$ represent the parametric line describing the stream in galactocentric coordinates, $\tau$ is the stream coordinate in time as described in the previous section, and is used as the independent variable in this analysis. The coefficients $a_i$, $b_i$, and $c_i$ are the polynomial coefficients. We found that a second-order polynomial was insufficient to capture the curvature along the full length of the stream, with divergence from the star particle positions occurring at the extremes. A third-order polynomial was sufficient and desirable, as it is the lowest order that adequately captures the path of the stream over the entire length under consideration.

@@

    As a result, each polynomial coefficient can now be expressed as a function of time. Consequently, we can parameterize the stream as a function of both simulation time and position along the stream:
    \begin{equation}
      \vec{s}(t,\tau) = 
      \left\{
      \begin{aligned}
        x(t,\tau) &= a_0(t) + a_1(t)\tau + a_2(t) \tau^2 + a_3(t)\tau^3 \\ 
        y(t,\tau) &= b_0(t) + b_1(t)\tau + b_2(t) \tau^2 + b_3(t)\tau^3 \\
        z(t,\tau) &= c_0(t) + c_1(t)\tau + c_2(t) \tau^2 + c_3(t)\tau^3.
        \end{aligned}
      \right.
    \end{equation}
    The values of the coefficients as a function of time are obtained through linear interpolation, ensuring that the coefficients at the snapshot times match the values constrained by the simulation data.

    Next, we fit the trajectory of the perturber with a second-order polynomial. With equations for both the stream and the perturber as functions of time, we identify the time and location of impact by minimizing a cost function, defined as the distance between the stream and the perturber:
    \begin{equation} 
      b(t, \tau) = \left\lVert \vec{s}(t, \tau) - \vec{p}(t) \right\rVert, 
      \end{equation}
    where $\vec{s}(t, \tau)$ is the galactocentric position of a point on the stream, $\vec{p}(t)$ is the position of the perturber, and $b$ is the distance between the two. The minimum value of $b$, denoted as $\text{min}(b)$, represents the impact parameter. The minimization is carried out using \texttt{scipy}'s optimization package with the \textit{L-BFGS-B} method, which allows us to constrain $t$ and $\tau$, ensuring no extrapolation occurs.

    Once this minimization is performed, determining the relative velocity becomes straightforward. Since the minimization provides the impact parameter, time of impact, and corresponding value of $\tau$, we can compute the derivatives of the parametric equations at $t_{\text{min}}$ and $\tau_{\text{min}}$. The parallel and perpendicular components of the perturber's velocity relative to the stream are given by:
    \begin{equation}
      \begin{aligned}
        \delta \vec{v} &=\vec{v_p} - \vec{v_s} \\
        w_\parallel &= \left(\delta \vec{v}\right)\cdot \hat{v_s}\\  
        w_\perp &=  \sqrt{\Delta v ^2 - w_\parallel ^ 2}
        \end{aligned}
      \end{equation}
    where $\vec{v}_p$ and $\vec{v}_s$ are the velocities of the perturber and the stream, respectively. For each of the 50 simulations, this information was compute for the strongs 5 flybys of a perturber with the stream. Thus, we compute a sample of 250 impacts, and flag those that give way to gaps.
